---
title: "ISMRM25"
author: "Francesco Santini"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("pacman")) install.packages("pacman")
pacman::p_load('tidyverse', 'gridExtra', 'ggplot2', 'patchwork', 'dplyr', 'tidyr')

set.seed(123)
```

## Distributions

```{r distributions}
# R script to create violin plots of two normal distributions
# Parameters can be adjusted as needed

# Set the sample size (user-definable variable)
sample_size <- 1000  # Change this value to adjust the number of samples

# Generate normally distributed data
group_a <- rnorm(n = sample_size, mean = 1000, sd = 50)
group_b <- rnorm(n = sample_size, mean = 1005, sd = 50)

# Combine data for plotting
data <- data.frame(
  value = c(group_a, group_b),
  group = factor(rep(c("Measurement 1", "Measurement 2"), each = sample_size))
)

# Create violin plots
plot <- ggplot(data, aes(x = group, y = value, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red") +
  scale_fill_manual(values = c("dodgerblue", "firebrick")) +
  labs(
    title = paste("Violin Plots (n =", sample_size, ")"),
    x = "Group",
    y = "Value",
    fill = "Distribution"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

t.test(group_a, group_b)
mean(group_a)
mean(group_b)

plot
```

## The p-value


```{r p_value_vis}

## Parameters

value <- 1.5

##


# Create normal distribution data
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)

# Create data frame for plotting
df <- data.frame(x = x, y = y)

p_value <- (1-pnorm(value))

# Create the plot
ggplot(df, aes(x = x, y = y)) +
  geom_line() +
  geom_area(data = subset(df, x >= value), fill = "red", alpha = 0.3) +
  labs(x = "Test Statistic", y = "Density",
       title = paste("Normal Distribution with the p-value for",
                     value, "represented as area (p=", round(p_value, 3), ")")) +
  theme_minimal()
```

# Confidence intervals

The t-distribution can be used to calculate *confidence intervals* for our means. A X% confidence interval comes from a procedure that, if repeated many times with different random samples, would contain the true population mean X% of the time. When talking about means, we can use the t-distribution to build a confidence interval in the following way:

$$CI(X\%) = \bar x \pm t(X\%) \cdot (\hat\sigma/\sqrt n)$$
where $t(X\%)$ is the critical value that corresponds to the desired confidence percentage. In formulas:

```{r confidence interval}

## Parameters

population_mean <- 4
population_sd <- 1
sample_size <- 10
confidence <- 0.95

##

standard_error <- population_sd / sqrt(sample_size)

confidence_onesided <- 1 - (1-confidence)/2
t_crit <- qt(confidence_onesided, df=sample_size-1)

# Create data frame for plotting
x <- seq(population_mean - 4*standard_error, population_mean + 4*standard_error, length.out = 1000)
y <- dt((x - population_mean) / standard_error, df=sample_size-1)

ci_min <- population_mean - t_crit * standard_error
ci_max <- population_mean + t_crit * standard_error
```

## Table of coefficients to calculate the 95% C.I.

```{r CI table}

confidence <- 0.95
confidence_onesided <- 1- (1-confidence)/2


se_mult_factor <- function(N) qt(confidence_onesided, df=N-1)
sd_mult_factor <- function(N) se_mult_factor(N)/sqrt(N)

n_array <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 30, 50, 100, 1000)
se_mult_factor_array <- sapply(n_array, se_mult_factor)
sd_mult_factor_array <- sapply(n_array, sd_mult_factor)


table_df <- data.frame(
  SampleSize = n_array,
  SE_Multiplier =  se_mult_factor_array,
  SD_Multiplier = sd_mult_factor_array
)

knitr::kable(table_df, col.names = c("Sample Size", "t(97.5%)", "Multiplier for the SD"))
```

## Table of coefficients for the C.I. of the SD
```{r CI table SD}

confidence <- 0.95

alpha <- 1-confidence
lower_limit <- function(N) sqrt((N-1) / qchisq(1-(alpha/2), df = N-1))
upper_limit <- function(N) sqrt((N-1) / qchisq(alpha/2, df = N-1))

n_array <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 30, 50, 100, 1000)
lower_limit_array <- sapply(n_array, lower_limit)
upper_limit_array <- sapply(n_array, upper_limit)


table_df <- data.frame(
  SampleSize = n_array,
  LL =  lower_limit_array,
  UL = upper_limit_array
)

knitr::kable(table_df, col.names = c("Sample Size", "Lower limit multiplier", "Upper limit multiplier"))
```

## Bland Altman with confidence intervals

```{r BA with CI}
# Function to create Bland-Altman plot with CI adjustments for sample size using ggplot2
bland_altman_with_ci <- function(method1, method2, alpha = 0.05) {
  # Calculate differences and means
  differences <- method1 - method2
  means <- (method1 + method2) / 2
  n <- length(differences)
  
  # Mean difference (bias)
  mean_diff <- mean(differences)
  
  # Standard deviation of differences
  sd_diff <- sd(differences)
  
  # Standard error of the mean difference
  se_mean <- sd_diff / sqrt(n)
  
  # t-critical value for the given alpha and df
  t_crit <- qt(1 - alpha/2, df = n - 1)
  
  # Confidence interval for the mean difference
  mean_ci_lower <- mean_diff - t_crit * se_mean
  mean_ci_upper <- mean_diff + t_crit * se_mean
  
  # Standard limits of agreement
  lower_loa <- mean_diff - 1.96 * sd_diff
  upper_loa <- mean_diff + 1.96 * sd_diff
  
  # Standard error of the limits of agreement
  # Formula from Bland & Altman (1999): se_limits = sqrt((3*sd_diff^2)/n)
  se_limits <- sqrt((3 * sd_diff^2) / n)
  
  # Confidence intervals for the limits of agreement
  lower_loa_ci_lower <- lower_loa - t_crit * se_limits
  lower_loa_ci_upper <- lower_loa + t_crit * se_limits
  upper_loa_ci_lower <- upper_loa - t_crit * se_limits
  upper_loa_ci_upper <- upper_loa + t_crit * se_limits
  
  # Create a data frame for plotting
  plot_data <- data.frame(means = means, differences = differences)
  
  # Find the maximum absolute value for symmetric y-axis limits
  y_limit <- max(abs(c(
    lower_loa_ci_lower, upper_loa_ci_upper,
    min(differences), max(differences)
  ))) * 1.1  # Add 10% padding
  
  # Create the plot with ggplot2
  p <- ggplot(plot_data, aes(x = means, y = differences)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
    geom_hline(yintercept = mean_diff, linetype = "solid", color = "blue", linewidth = 1) +
    geom_hline(yintercept = lower_loa, linetype = "dashed", color = "red", linewidth = 0.8) +
    geom_hline(yintercept = upper_loa, linetype = "dashed", color = "red", linewidth = 0.8) +
    # Add confidence intervals for the mean difference
    geom_hline(yintercept = mean_ci_lower, linetype = "dotted", color = "blue", linewidth = 0.8) +
    geom_hline(yintercept = mean_ci_upper, linetype = "dotted", color = "blue", linewidth = 0.8) +
    # Add confidence intervals for the limits of agreement
    geom_hline(yintercept = lower_loa_ci_lower, linetype = "dotted", color = "red", linewidth = 0.8) +
    geom_hline(yintercept = lower_loa_ci_upper, linetype = "dotted", color = "red", linewidth = 0.8) +
    geom_hline(yintercept = upper_loa_ci_lower, linetype = "dotted", color = "red", linewidth = 0.8) +
    geom_hline(yintercept = upper_loa_ci_upper, linetype = "dotted", color = "red", linewidth = 0.8) +
    # Labels and title
    labs(
      title = paste0("Bland-Altman Plot (n = ", n, ")"),
      subtitle = "With 95% Confidence Intervals",
      x = "Mean of Methods",
      y = "Difference (Method 1 - Method 2)"
    ) +
    # Set symmetric y-axis limits
    ylim(-y_limit, y_limit) +
    # Add a legend
    annotate("text", x = max(means), y = mean_diff, 
             label = paste("Mean =", round(mean_diff, 2)), 
             hjust = 1, vjust = -0.5, color = "blue") +
    annotate("text", x = max(means), y = upper_loa, 
             label = paste("Upper LoA =", round(upper_loa, 2)), 
             hjust = 1, vjust = -0.5, color = "red") +
    annotate("text", x = max(means), y = lower_loa, 
             label = paste("Lower LoA =", round(lower_loa, 2)), 
             hjust = 1, vjust = 1.5, color = "red") +
    # Theme adjustments
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
  
  # Add custom legend
  p <- p + 
    annotate("segment", x = min(means), y = -y_limit*0.9, xend = min(means) + (max(means)-min(means))*0.05, 
             yend = -y_limit*0.9, color = "blue", linewidth = 1) +
    annotate("text", x = min(means) + (max(means)-min(means))*0.06, y = -y_limit*0.9, 
             label = "Mean difference", hjust = 0) +
    
    annotate("segment", x = min(means) + (max(means)-min(means))*0.25, y = -y_limit*0.9, 
             xend = min(means) + (max(means)-min(means))*0.3, yend = -y_limit*0.9, 
             color = "blue", linetype = "dotted", linewidth = 0.8) +
    annotate("text", x = min(means) + (max(means)-min(means))*0.31, y = -y_limit*0.9, 
             label = "Mean difference 95% CI", hjust = 0) +
    
    annotate("segment", x = min(means) + (max(means)-min(means))*0.6, y = -y_limit*0.9, 
             xend = min(means) + (max(means)-min(means))*0.65, yend = -y_limit*0.9, 
             color = "red", linetype = "dashed", linewidth = 0.8) +
    annotate("text", x = min(means) + (max(means)-min(means))*0.66, y = -y_limit*0.9, 
             label = "Limits of agreement", hjust = 0) +
    
    annotate("segment", x = min(means), y = -y_limit*0.8, 
             xend = min(means) + (max(means)-min(means))*0.05, yend = -y_limit*0.8, 
             color = "red", linetype = "dotted", linewidth = 0.8) +
    annotate("text", x = min(means) + (max(means)-min(means))*0.06, y = -y_limit*0.8, 
             label = "Limits of agreement 95% CI", hjust = 0)
  
  # Print a summary of the results
  cat("Sample size:", n, "\n")
  cat("Mean difference:", round(mean_diff, 2), "with 95% CI [", 
      round(mean_ci_lower, 2), ",", round(mean_ci_upper, 2), "]\n")
  cat("Limits of agreement:\n")
  cat("  Lower:", round(lower_loa, 2), "with 95% CI [", 
      round(lower_loa_ci_lower, 2), ",", round(lower_loa_ci_upper, 2), "]\n")
  cat("  Upper:", round(upper_loa, 2), "with 95% CI [", 
      round(upper_loa_ci_lower, 2), ",", round(upper_loa_ci_upper, 2), "]\n")
  
  # Return both the plot and the numeric results
  return(list(
    plot = p,
    results = list(
      n = n,
      mean_diff = mean_diff,
      sd_diff = sd_diff,
      mean_ci_lower = mean_ci_lower,
      mean_ci_upper = mean_ci_upper,
      lower_loa = lower_loa,
      upper_loa = upper_loa,
      lower_loa_ci_lower = lower_loa_ci_lower,
      lower_loa_ci_upper = lower_loa_ci_upper,
      upper_loa_ci_lower = upper_loa_ci_lower,
      upper_loa_ci_upper = upper_loa_ci_upper
    )
  ))
}

# Function to run simulation with different sample sizes
simulate_ba_plots <- function(n) {
  set.seed(123)  # For reproducibility
  
  # True values with some random error
  true_values <- rnorm(n, mean = 100, sd = 15)
  
  # Method 1: introduces a systematic bias and random error
  method1 <- true_values + rnorm(n, mean = 5, sd = 8)
  
  # Method 2: introduces only random error
  method2 <- true_values + rnorm(n, mean = 0, sd = 8)
  
  # Create Bland-Altman plot with CI
  result <- bland_altman_with_ci(method1, method2)
  result$plot
}

simulate_ba_plots(5)
simulate_ba_plots(10)
simulate_ba_plots(20)
simulate_ba_plots(50)
```

## Plot of a normal distribution and distribution of differences

```{r normdist}

# Set seed for reproducibility
set.seed(123)

# Parameters
n_samples <- 10000  # Number of samples to generate
mean_val <- 0       # Mean of the normal distribution
sd_val <- 1         # Standard deviation of the normal distribution

# Generate samples from normal distribution
samples <- rnorm(n_samples, mean = mean_val, sd = sd_val)

# Generate another set of samples
samples2 <- rnorm(n_samples, mean = mean_val, sd = sd_val)

# Calculate differences between random pairs of samples
differences <- samples - samples2

# Create a data frame for plotting
plot_data <- data.frame(
  value = c(samples, differences),
  type = factor(c(rep("Original Samples", n_samples), 
                  rep("Differences", n_samples)),
                levels = c("Original Samples", "Differences"))
)

# Calculate theoretical values for plotting the normal distribution curves
x_range <- seq(min(plot_data$value), max(plot_data$value), length.out = 1000)
original_density <- dnorm(x_range, mean = mean_val, sd = sd_val)
diff_density <- dnorm(x_range, mean = 0, sd = sqrt(2)*sd_val)

curve_data <- data.frame(
  x = c(x_range, x_range),
  y = c(original_density, diff_density),
  type = factor(c(rep("Original Samples", 1000), 
                  rep("Differences", 1000)),
                levels = c("Original Samples", "Differences"))
)

# Calculate statistics for display
orig_mean <- mean(samples)
orig_sd <- sd(samples)
diff_mean <- mean(differences)
diff_sd <- sd(differences)


# Create the plot
p <- ggplot() +
  # Add histograms (normalized to form density)
  geom_histogram(data = plot_data, 
                 aes(x = value, fill = type, color = type),
                 alpha = 0.5, position = "identity",
                 binwidth = 0.2) +
  
  # Add theoretical density curves
  geom_line(data = curve_data,
            aes(x = x, y = y, color = type),
            size = 1) +
  
  # Customize colors
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_color_manual(values = c("steelblue4", "firebrick4")) +
  
  
  labs(
    title = "Comparison of Normal Distribution and Differences Distribution",
    subtitle = paste("Original: N(", mean_val, ",", sd_val, "), Differences: N(0, 2*", sd_val, ")"),
    x = "Value",
    y = "Density",
    fill = "Distribution",
    color = "Theoretical Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
p
```
